Implementing simplified INLA into `aghq`

### 2022-10-19 src/dev_sinla/sinla.Rmd
Add CDFs, KS tests, and checks on trapezoid rules working

```{r message=FALSE, warning=FALSE}
plot <- extract(mcmc, pars = "phi_prev[1]") %>%
  as.data.frame() %>%
  ggplot(aes(x = phi_prev.1.)) +
    geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7]) +
    theme_minimal() +
    labs(x = "phi_prev[1]", y = "Posterior")

plot
```

@@

#' Hessian evaluated at the mean (or equivalently mode)
Q <- h$env$spHess(mm, random = TRUE)
dim(Q) #' With dimension [30 x 30]

cowplot::plot_grid(plot(mean), image(Q), nrow = 1)
```

@@

Now let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan` we obtained in Section \@ref(ground-truth).
As described above, we simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
$$
\begin{align}
\tilde p(x_i \, | \, y) &= \tilde p_\text{G}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
&\propto \tilde p_\text{G}(x_i \, | \, \hat \theta, y)
\end{align}
$$
We can evaluate this for $i = 1$ via:

@@

And compare it to our MCMC results as follows:

```{r message=FALSE, warning=FALSE}
plot2 <- plot +
  stat_function(
    data = data.frame(x = c(0.5, 3.5)), aes(x, col = "Joint Gaussian marginal"), fun = dnorm, n = 101, 
    args = list(mean = mean_i, sd = sqrt(var_i)), size = 1
  ) +
  labs(x = "prev_phi[1]", y = "Posterior", col = "Approximation") +
  scale_colour_manual(values = cbpalette) +
  theme_minimal()

plot2
```

Take some samples from the posterior marginal:

```{r}
M <- 1000
samples_jointgaussianmarginal <- rnorm(M, mean_i, sqrt(var_i)) 
```

@@

@@

  #' Starting parameters for TMB are the same for every index of the loop
  param_i <- list(
    beta_prev = 0,
    phi_prev_i = 0,
    phi_prev_minus_i = rep(0, data$n - 1),
    log_sigma_phi_prev = 0
  )

@@

Q <- Qs[[i]]
dim(Q) #' With dimension [(30 - 1) x (30 - 1)]

cowplot::plot_grid(plot(mean), image(Q))
```

`obj$fn` should take as input $(x_i, \theta)$:

@@

df <- opt$par %>%
  t() %>%
  as.data.frame() %>%
  slice(rep(1:n(), each = x_grid_length)) %>%
  mutate(
    phi_prev_i = x_grid
  ) %>%
  rowwise() %>%
  mutate(
    #' Note here: it's crucial that the arguments are in the right order!
    logpost = -obj$fn(c(beta_prev, phi_prev_i, log_sigma_phi_prev))
  )

ggplot(df, aes(x = phi_prev_i, y = logpost)) +
  geom_line() +
  theme_minimal() +
  labs(x = "phi_prev[1]", "log(posterior)")
```

What we've got here is the unnormalised log-posterior.
How can we go from that to a normalised posterior?
Going to do trapezoid rule here, but AGHQ would work better (even with a low number of points) because the function is close to being Gaussian.

@@

@@

@@

```{r}
max(abs(df$post_norm - df$post_norm_alt))
```

@@

```{r message=FALSE, warning=FALSE}
plot3 <- plot2 +
  geom_line(data = df, aes(x = phi_prev_i, y = post_norm, col = "Laplace"), size = 1)

plot3
```

```{r}
samples_tmbstan <- extract(mcmc, pars = "phi_prev[1]")[[1]]
ks_jointgaussianmarginal_tmbstan <- inf.utils::ks_test(samples_tmbstan, samples_jointgaussianmarginal)
```

One thing to take into account is that we're doing empirical Bayes, and things might be improved by using more integration points.

>>>>

```{r message=FALSE, warning=FALSE}
samples_tmbstan <- extract(mcmc, pars = "phi_prev[1]")[[1]]

plot_pdf <- ggplot(data.frame(x = samples_tmbstan), aes(x = x)) +
    geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7]) +
    theme_minimal() +
    labs(x = "phi_prev[1]", y = "Posterior PDF")

ecdf_tmbstan <- ecdf(samples_tmbstan)
grid <- seq(min(samples_tmbstan), max(samples_tmbstan), by = 0.1)

plot_cdf <- data.frame(x = grid, y = ecdf_tmbstan(grid)) %>%
  ggplot(aes(x = x, y = y)) +
    geom_line(col = cbpalette[7]) +
    theme_minimal() +
    labs(x = "phi_prev[1]", y = "Posterior CDF")

cowplot::plot_grid(plot_cdf, plot_pdf)
```

@@

#' Hessian evaluated at the mean (or equivalently mode)
Q <- h$env$spHess(mm, random = TRUE)
dim(Q) #' With dimension [30 x 30]
```

@@

Now let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan` we obtained in Section \@ref(ground-truth).
As described above, we simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
$$
\begin{align}
\tilde p(x_i \, | \, y) &= \tilde p_\text{G}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
&\propto \tilde p_\text{G}(x_i \, | \, \hat \theta, y)
\end{align}
$$

We can evaluate this for $i = 1$ via:

@@

Because this is a Gaussian distribution, we do not have do normalise it (as would usually be the case) because we know it's a density.
We can compare it to our MCMC results as follows:

```{r message=FALSE, warning=FALSE}
plot_pdf2 <- plot_pdf +
  stat_function(
    data = data.frame(x = c(0.5, 3.5)), aes(x, col = "Joint Gaussian marginal"), fun = dnorm, n = 101, 
    args = list(mean = mean_i, sd = sqrt(var_i)), size = 1
  ) +
  labs(x = "prev_phi[1]", y = "Posterior", col = "Approximation") +
  scale_colour_manual(values = cbpalette) +
  theme_minimal()

plot_pdf2

M <- 1000
samples_jointgaussianmarginal <- rnorm(M, mean_i, sqrt(var_i))

ecdf_jointgaussianmarginal <- ecdf(samples_jointgaussianmarginal)
grid <- seq(min(samples_jointgaussianmarginal), max(samples_jointgaussianmarginal), by = 0.1)

plot_cdf2 <- plot_cdf +
  geom_line(
    data = data.frame(x = grid, y = ecdf_jointgaussianmarginal(grid)),
    aes(x = x, y = y), col = cbpalette[1]
  )

cowplot::plot_grid(plot_cdf2, plot_pdf2, rel_widths = c(1, 1.5))
```

```{r}
ks_jointgaussianmarginal_tmbstan <- inf.utils::ks_test(samples_tmbstan, samples_jointgaussianmarginal)
ks_jointgaussianmarginal_tmbstan
```

The maximum difference between the two ECDFS is `r ks_jointgaussianmarginal_tmbstan$D` which occurs at the point `r ks_jointgaussianmarginal_tmbstan$l`.

@@

#' Starting parameters for TMB are the same for every index of the loop so it's more efficient to put them outside
param_i <- list(
  beta_prev = 0,
  phi_prev_i = 0,
  phi_prev_minus_i = rep(0, data$n - 1),
  log_sigma_phi_prev = 0
)

@@

@@

Q <- Qs[[i]]
dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
```

We check here that `obj$fn` indeed takes as input $(x_i, \theta)$:

@@

df <- opt$par %>%
  t() %>%
  as.data.frame() %>%
  slice(rep(1:n(), each = x_grid_length)) %>%
  mutate(
    phi_prev_i = x_grid
  ) %>%
  rowwise() %>%
  mutate(
    #' Note here: it's crucial that the arguments are in the right order!
    logpost = -obj$fn(c(beta_prev, phi_prev_i, log_sigma_phi_prev))
  )
```

We've compuating the unnormalised log-posterior -- how can we go from that to a normalised posterior?
Going to do trapezoid rule here, but AGHQ would work better (even with a low number of points) because the function is close to being Gaussian.

@@

Check that this method correctly computes the normalising constant for a density (which is known to integrate to one):

```{r}
abs(trapezoid_rule(dnorm(seq(-10, 10, by = 0.1)), spacing = 0.1) - 1) < 10E-6
```

@@

Check that this method correctly computes the normalising constant for a density (which is known to integrate to one):

```{r}
abs(trapezoid_rule_log(dnorm(seq(-10, 10, by = 0.1), log = TRUE), spacing = 0.1) - log(1)) < 10E-6
```

@@

```{r}
max(abs(df$post_norm - df$post_norm_alt)) < 10E-6
```

@@

```{r message=FALSE, warning=FALSE}
plot_pdf3 <- plot_pdf2 +
  geom_line(data = df, aes(x = phi_prev_i, y = post_norm, col = "Laplace"), size = 1)

plot_pdf3

df$ecdf <- cumsum(df$post_norm) * c(0, diff(df$phi_prev_i))

plot_cdf3 <- plot_cdf2 +
  geom_line(
    data = df, aes(x = phi_prev_i, y = ecdf), col = cbpalette[2]
  )

plot_cdf3

cowplot::plot_grid(plot_cdf3, plot_pdf3, rel_widths = c(1, 1.5))
```

We use the inverse CDF method to sample from the Laplace marginal, then we can calculate the KS statistic:

```{r}
samples_laplace <- approxfun(df$ecdf, df$phi_prev_i)(runif(M))

ks_laplace_tmbstan <- inf.utils::ks_test(samples_tmbstan, samples_laplace)

kable_data <- data.frame(
  "KS Distance" = c(ks_jointgaussianmarginal_tmbstan$D, ks_laplace_tmbstan$D),
  "KS Location" = c(ks_jointgaussianmarginal_tmbstan$l, ks_laplace_tmbstan$l)
)

rownames(kable_data) <- c("$\\text{KS}(\\texttt{tmbstan}, \\texttt{jointgaussianmarginal})$", "$\\text{KS}(\\texttt{tmbstan}, \\texttt{laplace})$")

kableExtra::kable(kable_data, booktabs = TRUE, escape = FALSE, align = "c")
```
